Predicting the effects of character displacement on food-web dynamics

### 2019-11-18 manuscript_ECD_model.tex
incorporate final (minor) edits to main and supplementary text.

\usepackage{fullpage}
\linespread{1.5}
\usepackage{lineno}
\usepackage{caption}
\captionsetup[figure]{font=small}

@@

\emph{Data accessibility}: No new data were used. All code used to
reproduce this manuscript and simulations within are publically
available on GitHub (\url{https://github.com/mabarbour/ECD_model}) and
have been archived with Zenodo (INSERT DOI).

@@

\emph{Running title}: Character divergence destabilizes food webs

@@

\emph{Manuscript type}: Letter.

\emph{Number of words in abstract}: 143; \emph{Number of words in main
text}: 4086; \emph{Number of words in text boxes}: NA; \emph{Number of
references}: 45; \emph{Number of figures}: 4; \emph{Number of tables}:
0; \emph{Number of text boxes}: 0

@@

Although models that included resources led to insights to the evolution
of character displacement, the ecological feedback onto
consumer-resource dynamics has received surprisingly little attention.
This is likely because the ecological feedback has been primarily
studied through the lens of coexistence theory
\citep{Lawlor1976, Germain2018, Bassar2017, McPeek2019}. For example,
early theoretical work showed that ecological character displacement
promotes coexistence by favoring specialized consumers that experience
reduced interspecific competition \citep{Lawlor1976}. Yet, this
reduction in interspecific competition may, at the same time, increase
interspecific interactions between specialized consumers and their
resources. Both food-web theory and empirical studies have shown that
increasing the strength of consumer-resource interactions often
suppresses the abundance of resources, which if sufficient enough, can
generate oscillations and less stable consumer-resource dynamics
\citep{Rosenzweig1971, Luckinbill1973, Murdoch2002, Murdoch2003, McCann2011}.
Thus, a food-web perspective, which accounts for both the direct and
indirect effects of consumer-resource interactions, may yield new
insight to the ecological consequences of character displacement.

Here, I address this knowledge gap by studying a mathematical model that
examines how ecological character displacement affects consumer-resource
dynamics in a food-web context. I address two questions: (1) How does
ecological character displacement affect resource abundances? (2) How
does character displacement affect food-web stability? To test the
generality of these effects, I explored different ecological foraging
scenarios and evolutionary tradeoffs in consumer attack rates. I found
that the adaptive process of character displacement often comes with an
ecological cost; resulting in food webs with lower resource availability
and that are less resilient to perturbations.

@@

\subsection{Underlying Consumer-Resource
Dynamics}\label{underlying-consumer-resource-dynamics}

@@

where \(r_i\) represents the intrinsic growth rate of resource \(i\),
\(K_i\) represents the carrying capacity of resource \(i\), \(e_{ij}\)
represents the conversion efficiency of resource \(i\) into consumer
\(j\), and \(m_j\) represents the mortality rate of consumer \(j\).
\(F_{ij}(R_i)\) represents consumer \(j\)'s feeding rate on resource
\(i\) (i.e.~functional response). This model is a useful
characterization of a scenario where consumers compete for two distinct
resources (e.g.~zooplankton and benthic invertebrates in lakes) rather
than a scenario where resources are better characterized by a continuous
trait distribution (e.g.~seed size, see \citet{Taper1985} for an
example). Importantly, inferences about ecological character
displacement can only be made by comparing food webs with and without a
competing consumer \citep{Schluter1992}. Therefore, I arbitrarily set
\(C_2=0\) to create a food-web without a competing consumer for these
comparisons.

\subsection{Foraging Scenarios}\label{foraging-scenarios}

I studied three different foraging scenarios. In the first, I assume
that consumers can forage for both resources simultaneously (fig.
\ref{fig:foraging_scenarios} top) and their feeding rates increase
linearly with resource abundance, such that:

@@

where \(a_{ij}\) is the attack rate of consumer \(j\) on resource \(i\).
This first scenario is the starting point for many models of resource
competition \citep{MacArthur1972}; however, it does not reflect many
food webs where consumers are mobile and their foraging behavior links
resources that occur in different habitats \citep{McCann2005}. The
second scenario accounts for this spatial context (fig.
\ref{fig:foraging_scenarios} bottom) and takes the form:

@@

where \(w_{ij}\) represents the proportion of time consumer \(j\) spends
foraging in a habitat where only resource \(i\) is found (i.e.~habitat
preference). Note that since \(w_{ij}\) is a proportion that
\(w_{1,j}=1-w_{2,j}\). Finally, it is well known that consumer feeding
rates often saturate at high resource abundances
\citep{Holling1959, Rosenzweig1963, Murdoch2003, McCann2011} and that
consumers do not usually spend a fixed proportion of time in a
particular habitat \citep{McCann2005}. The third scenario accounts for
these biological realities and takes the form \citep[derived
by][]{McCann2005}:\\

@@

where consumer \(j\)'s feeding rate on resource \(i\) is influenced by
the abundance of each resource; saturates as resource abundances
increase (due to handling times \(h_{ij}\)); and consumer habitat
preferences are modified by the relative abundance of resources, such
that: \(W_{ij}=\frac{w_{ij}R_i}{w_{1,j}R_1+w_{2,j}R_2}\).

Previous studies have analyzed the evolution of consumer attack rates in
the first two foraging scenarios using an Adaptive Dynamics approach,
with the general result being character divergence
\citep{Lawlor1976, Abrams1986}. I also used an Adaptive Dynamics
approach to analyze character displacement in the third foraging
scenario, and I too observed character divergence (detailed analysis in
Appendix S1 of Supplementary Information). I say consumers have
undergone character divergence if their evolved attack rates are more
specialized when evolving with vs.~without a competing consumer.
Specialization of consumer \(j\) on resource \(1\) is measured as
\(\frac{a_{1,j}}{a_{1,j}+a_{2,j}}\), where a value of 0.5 is a complete
generalist (\(a_{1,j}=a_{2,j}\)), and a value of 1 is a complete
specialist (\(a_{2,j}=0\)). Values less than 0.5 indicate specialization
on the other resource.

\begin{figure}
\centering
\includegraphics{Fig_1_ForagingScenarios}
\caption{\label{fig:foraging_scenarios}\textbf{Ecological foraging
scenarios.} I examined whether the effect of ecological character
displacement on food-web dynamics depended on whether consumers competed
for resources that occurred in the same vs.~different habitats. The
width of each arrow corresponds to initial attack rates
(\emph{a}\textsubscript{ij}) of consumer \emph{j} on resource \emph{i}.
Note that \emph{C}\textsubscript{1} was pre-adapted to
\emph{R}\textsubscript{1} (\emph{a}\textsubscript{11} \textgreater{}
\emph{a}\textsubscript{21}), while \emph{C}\textsubscript{2} was a
mirror image, being pre-adapted to \emph{R\textsubscript{2}}
(\emph{a}\textsubscript{22} \textgreater{} \emph{a}\textsubscript{12}).
In each scenario, I assumed consumer feeding rates increased linearly
with resource abundance. I also relax this assumption and consider a
more realistic functional response when resources occurred in different
habitats.}
\end{figure}

\subsection{Food-web Dynamics}\label{food-web-dynamics}

Given that character divergence occurred across these foraging
scenarios, I focus here on its consequences for food-web dynamics. To do
this, I analyzed differences in resource abundances and food-web
stability at equilibrium. An equilibrium is reached when the rates of
change in equation \ref{eq:1} are 0, and solving the system at this
point gives equilibrium abundances for each resource (\(\hat R_i\)) and
each consumer (\(\hat C_j\)). I also compared the local stability of
these food webs using standard methods \citep{Otto2007}. This stability
analysis derives the dominant eigenvalue, \(\lambda_{\text{max}}\), of
the matrix of partial derivatives of all differential equations with
respect to all variables evaluated at equilibrium. If
\(-\lambda_{max}>0\), then the food web will return to equilibrium after
a small perturbation (i.e.~locally stable), with more positive values
indicating a faster return time. If \(-\lambda_{max}<0\), then the food
web is not locally stable.

When possible, I derived analytical expressions for the relationship
between consumer attack rates and food-web dynamics. To do this, I
simplified the model by assuming that resources are equivalent
(\(r=r_i\) and \(K=K_i\)) as well as consumers (\(e=e_{ij}\);
\(h=h_{ij}\); \(m=m_j\)), except that consumer attack rates and their
habitat preferences (if present) are mirror images of each other
(\(a_{11}=a_{22}\); \(a_{12}=a_{21}\); \(w_{11}=w_{22}\)). Note that I
arbitrarily set \(C_1\) as being pre-adapted to \(R_1\)
(\(a_{11} > a_{21}\); \(w_{11} > 0.5\)), and therefore \(C_2\) was
pre-adapted to \(R_2\). Controlling for other sources of variability
allowed me to isolate the general effects of character divergence. All
mathematical derivations were conducted in Mathematica
\citep{Mathematica} and are provided in the Supplementary Information
(Appendix S1-3).

To gain insight to the eco-evolutionary feedback generated by character
displacement, I conducted simulations using an Adaptive Dynamics
approach. Specifically, after letting consumer and resource abundances
reach a steady state, I created a mutant consumer by randomly choosing
one and modifying its attack rate on one resource by either subtracting
or adding a small constant (0.01 in the following simulations) with
equal probability. The mutant's attack rate on the other resource was
determined by a tradeoff, such that
\(\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1\),
where \(A\) is the total investment in attack rates and \(n\) describes
the shape of the tradeoff \citep{Sargent2006}. This function has the
useful property that it differentiates between cases where intermediate
combinations of \(a_{\text{1},j}\) and \(a_{\text{2},j}\) are higher
than the extremes (when \(n>1\), green line in fig. \ref{fig:tradeoff})
or, conversely, where the two extremes are higher than intermediate
investments (when \(n<1\), orange line fig. \ref{fig:tradeoff}). When
\(n=1\), the tradeoff function is linear, and all combinations of
\(a_{\text{1},j}\) and \(a_{\text{2},j}\) have the same total attack
rate (blue line in fig. \ref{fig:tradeoff}). Assuming the mutant
consumer was rare, I then determined whether the mutant had higher
relative fitness than the resident consumer, and thus could invade and
replace the resident consumer. If the mutant was able to invade, I
updated the attack rate of the resident consumer to the mutant attack
rate and allowed consumer and resource abundances to reach a steady
state. I then repeated the simulation up to 10,000 times, which was
sufficient for consumers to either reach an evolutionary stable strategy
\citep[ESS,][]{Smith1973} or an evolutionary limit (e.g.
\(\frac{a_{ij}}{a_{\text{1},j}+a_{\text{2},j}}\) is constrained to a
maximum of 1 and minimum of 0). Unless otherwise noted, I conducted
simulations with the following parameter values: \emph{r} = 1; \emph{K}
= 4; \emph{e} = 0.8; \emph{m} = 1; \emph{A} = 2; \emph{h} = 0.4; and
\(w_{11} = w_{22} = 0.6\). I set an initial value of
\(a_{11} = a_{22} = 1.2\), while \(a_{12}\) and \(a_{21}\) depended on
the value of \emph{n}. I set initial consumer and resource abundances
to: \(R_1 = R_2 = 2\); \(C_1 = C_2 = 1\). All simulations were conducted
in R \citep{R} and the code to reproduce these simulations is publically
available on GitHub (\url{https://github.com/mabarbour/ECD_model}) and
have been archived with Zenodo (INSERT DOI).

\begin{figure}
\centering
\includegraphics{Fig_2_Tradeoffs.pdf}
\caption{\label{fig:tradeoff}\textbf{Evolutionary tradeoffs in consumer
attack rates.} In each foraging scenario, I explored the effects of
three different tradeoffs: intermediate combinations of attack rates
(\emph{a}\textsubscript{1,\emph{j}}, \emph{a}\textsubscript{2,\emph{j}})
are higher than the extremes (green line, \emph{n} \textgreater{} 1);
extreme combinations of attack rates are higher than intermediate
investments (orange line, \emph{n} \textless{} 1); and all combinations
of attack rates have the same total attack rate (blue line, \emph{n} =
1). Points corresponding to attack rates at the beginning of the
simulation for \emph{C}\textsubscript{1}, which was pre-adapted to
\emph{R}\textsubscript{1} (\emph{a}\textsubscript{11} \textgreater{}
\emph{a}\textsubscript{12}).}
\end{figure}

@@

\subsection{Resources occur in same
habitat}\label{resources-occur-in-same-habitat}

@@

I found that the shape of the tradeoff function qualitatively affects
the relationship between character displacement and resource abundances
in this scenario (fig. \ref{fig:plot_fig3} middle left). For example, if
consumer's are constrained by a linear tradeoff (blue lines), then there
is no net change in total attack rate (fig. \ref{fig:plot_fig3} top
left) and character displacement has no effect on resource abundances
(fig. \ref{fig:plot_fig3} middle left). If the tradeoff is concave down
(green lines), then resource abundances can actually increase under
character displacement (fig. \ref{fig:plot_fig3} middle left). This is
because the total attack rate of consumers is maximized at intermediate
values (\(a_{1,j}=a_{2,j}\)) and decreases as consumers diverge (fig.
\ref{fig:plot_fig3} top left). When the tradeoff is concave up (orange
lines), character displacement suppresses resource abundances due to the
increase in total attack rates (fig. \ref{fig:plot_fig3} top left).
Although the equation I derived for resource abundances was for the
scenario where both consumers and both resources were present, it
accurately predicts the abundance of resources when a single consumer
reaches its evolutionary stable strategy (ESS; triangles on respective
colored lines in fig. \ref{fig:plot_fig3} middle left). This is because
a single consumer evolves to be a generalist that has equal attack rates
on each resource (triangles at 0.5 along x-axis in fig.
\ref{fig:plot_fig3} left panels), resulting in equivalent resource
abundances.

The effect of character displacement on resources corresponds to its
impact on food-web stability. For example, when character displacement
decreases resource abundances (orange points in fig. \ref{fig:plot_fig3}
middle left), there is a decrease in food-web stability (fig.
\ref{fig:plot_fig3} bottom left). Character divergence may not affect or
even increase food-web stability (blue and green lines in fig.
\ref{fig:plot_fig3} bottom left); however, evolution does not favor
strong divergence in these scenarios (blue and green points in fig.
\ref{fig:plot_fig3}), which dampens these contingent effects. Note that
the dip in stability occurs when both consumers evolve to be
generalists, a situation that is not favored in any of the foraging
scenarios we examined (fig. \ref{fig:plot_fig3} bottom).

@@

This equation implies that if consumers evolve to become specialists on
resources that occur in their preferred habitat (e.g. \(w_{1,j}>0.5\)
and \(a_{1,j}>a_{2,j}\)), then the effective attack rate of consumers
(\(w_{1,j}a_{1,j}+w_{2,j}a_{2,j}\)) will always increase, regardless of
the tradeoff (fig. \ref{fig:plot_fig3} top right). Thus, character
divergence always results in resource suppression (fig.
\ref{fig:plot_fig3} middle right). Note that the shape of the tradeoff
can modify the effect of character displacement. This is not so much due
to the tradeoff affecting the magnitude of displacement (it does, but
the effect is minor), but because the form of the tradeoff affects
resource abundances when a single consumer has reached an ESS (triangles
in fig. \ref{fig:plot_fig3} middle right). In contrast, resource
abundances reach a similar value when consumers evolve in the presence
of a competitor (circles in fig. \ref{fig:plot_fig3} middle right),
because character displacement tends to reach a constraint of complete
specialization. It is worth noting that resource abundances are
consistently higher at the single consumer ESS compared to the
predictions I derived for when both consumers are present (deviation of
triangles from respective colored lines in fig. \ref{fig:plot_fig3}
middle right). This is because consumers actually evolve to be slightly
specialized on the resources that occur in their non-preferred habitat
(deviation of triangles from 0.5 along x-axis in fig.
\ref{fig:plot_fig3} right panels).

As seen previously, the effect of character displacement on resource
abundances qualitatively corresponds to its effect on food-web stability
(fig. \ref{fig:plot_fig3}). Specifically, character divergence decreases
food-web stability, regardless of the tradeoff in attack rates. This is
not simply a consequence of having an additional consumer in the system,
but emerges from the eco-evolutionary feedback between character
displacement and resource suppression (fig. \ref{fig:plot_fig3}). For
example, when the tradeoff is concave up (orange), the initial
two-consumer food web (small circle) is more stable than when there is
only one consumer (small triangle); however, this pattern switches by
the end of the eco-evolutionary simulation (large points).

\begin{figure}
\centering
\includegraphics{Fig_3_MacArthur_LawlorSmith.pdf}
\caption{\label{fig:plot_fig3}\textbf{Effect of character displacement
on food-web dynamics under different evolutionary tradeoffs and foraging
scenarios.} Lines show predicted values when both consumers and
resources are present. Different line colors correspond to different
tradeoffs in attack rates (green, \emph{n} = 1.15; blue, \emph{n} = 1;
orange, \emph{n} = 0.85). Large circles (two consumers) and triangles
(one consumer) correspond to the end points of the eco-evolutionary
simulation for \emph{C}\textsubscript{1}, whereas as small shapes
correspond to the starting points (only in stability panels). In both
foraging scenarios, feeding rates increase linearly with resource
abundance, but the equation for effective attack rate is different.}
\end{figure}

@@

In the third foraging scenario (equation \ref{eq:4}), I observed the
same general effect of character divergence as the previous scenario
(resources in different habitats, but linear functional response). This
is because resource abundances at equilibrium are governed by a similar
dynamic (derived in Appendix S1):

@@

And since evolution favors consumer divergence onto their preferred
resources (see Appendix S1), the effective attack rate of consumers
(\(w_{1,j}a_{1,j}+w_{2,j}a_{2,j}\)) will always increase, resulting in
lower resource abundances and decreased food-web stability (Appendix
S4).

@@

\subsection{Resource Abundances}\label{resource-abundances}

>>>>

\usepackage{fullpage}
\linespread{1.5}
\usepackage{lineno}
\usepackage{caption,setspace}
\captionsetup[figure]{font=small}

@@

\emph{Data accessibility}: No new data were used. All code used to
reproduce this manuscript and simulations within are publically
available on GitHub (\url{https://github.com/mabarbour/ECD_model}).

@@

\emph{Running title}: Character displacement destabilizes food webs

@@

@@

Although models that included resources led to insights about the
evolution of character displacement, the ecological feedback onto
consumer-resource dynamics has received surprisingly little attention.
This is likely because the ecological feedback has been primarily
studied through the lens of coexistence theory
\citep{Lawlor1976, Germain2018, Bassar2017, McPeek2019}. For example,
early theoretical work showed that character displacement promotes
coexistence by favoring specialized consumers that experience reduced
interspecific competition \citep{Lawlor1976}. Yet, this reduction in
interspecific competition may, at the same time, increase interspecific
interactions between specialized consumers and their resources. Both
food-web theory and empirical studies have shown that increasing the
strength of consumer-resource interactions often suppresses the
abundance of resources, which if sufficient enough, can generate
oscillations and less stable consumer-resource dynamics
\citep{Rosenzweig1971, Luckinbill1973, Murdoch2002, Murdoch2003, McCann2011}.
Thus, a food-web perspective, which accounts for both the direct and
indirect effects of consumer-resource interactions, may yield new
insight to the ecological consequences of character displacement.

Here, I address this knowledge gap by studying a mathematical model that
examines how ecological character displacement affects consumer-resource
dynamics in a food-web context. Specifically, I sought to answer the
question: how does character displacement in consumer attack rates
affect resource abundances and food-web stability? To test the
generality of these effects, I explored different ecological foraging
scenarios and evolutionary tradeoffs in consumer attack rates. I found
that the adaptive process of character displacement often comes with an
ecological cost; resulting in food webs with lower resource availability
and that are less resilient to perturbations.

@@

\subsection{Underlying consumer-resource
dynamics}\label{underlying-consumer-resource-dynamics}

@@

where \(r_i\) represents the intrinsic growth rate of resource \(i\),
\(K_i\) represents the carrying capacity of resource \(i\), \(e_{ij}\)
represents the conversion efficiency of resource \(i\) into consumer
\(j\), and \(m_j\) represents the mortality rate of consumer \(j\).
\(F_{ij}(R_i)\) represents consumer \(j\)'s feeding rate on resource
\(i\) (i.e., its functional response). This model is a useful
characterization of a scenario where consumers compete for two distinct
resources (e.g.~zooplankton and benthic invertebrates in lakes) rather
than a scenario where resources are better characterized by a continuous
trait distribution (e.g., seed size; see \citet{Taper1985} for an
example). Importantly, inferences about character displacement can only
be made by comparing food webs with and without a competing consumer
\citep{Schluter1992}. Therefore, I arbitrarily set \(C_2=0\) to create a
food-web without a competing consumer for these comparisons.

\subsection{Foraging scenarios}\label{foraging-scenarios}

I studied three different foraging scenarios. In the first, I assume
that consumers can forage for both resources simultaneously (Fig.
\ref{fig:foraging_scenarios}a) and their feeding rate increases linearly
with resource abundance, such that:

@@

where \(a_{ij}\) is the attack rate of consumer \(j\) on resource \(i\).
This first scenario is the starting point for many models of resource
competition \citep{MacArthur1972}; however, it does not reflect many
food webs where consumers are mobile and their foraging behavior links
resources that occur in different habitats \citep{McCann2005}. The
second scenario accounts for this spatial context (Fig.
\ref{fig:foraging_scenarios}b) and takes the form:

@@

where \(w_{ij}\) represents the proportion of time consumer \(j\) spends
foraging in a habitat where only resource \(i\) is found (i.e., its
habitat preference). Note that since \(w_{ij}\) is a proportion that
\(w_{1,j}=1-w_{2,j}\). Finally, it is well known that consumer feeding
rates often saturate at high resource abundances
\citep{Holling1959, Rosenzweig1963, Murdoch2003, McCann2011} and that
consumers do not usually spend a fixed proportion of time in a
particular habitat \citep{McCann2005}. The third scenario accounts for
these biological realities and takes the form \citep[derived
by][]{McCann2005}:\\

@@

where consumer \(j\)'s feeding rate on resource \(i\) is influenced by
the abundance of each resource; saturates as resource abundances
increase (due to handling time \(h_{ij}\)); and consumer habitat
preferences are modified by the relative abundance of resources, such
that: \(W_{ij}=\frac{w_{ij}R_i}{w_{1,j}R_1+w_{2,j}R_2}\).

Previous studies have analyzed the evolution of consumer attack rates in
the first two foraging scenarios using an Adaptive Dynamics approach,
with the general result being divergent character displacement
\citep{Lawlor1976, Abrams1986}. I also used an Adaptive Dynamics
approach to analyze the evolution of consumer attack rates in the third
foraging scenario, and I too observed divergent character displacement
(detailed analysis given in Appendix S1). I say consumers have undergone
divergent character displacement if their evolved attack rates are more
specialized when evolving with vs.~without a competing consumer.
Specialization of consumer \(j\) on resource \(1\) is measured as
\(\frac{a_{1,j}}{a_{1,j}+a_{2,j}}\), where a value of 0.5 is a complete
generalist (\(a_{1,j}=a_{2,j}\)), and a value of 1 is a complete
specialist (\(a_{2,j}=0\)). Values less than 0.5 indicate specialization
on the other resource. Since I did not observe convergent character
displacement in any of the foraging scenarios I analyzed, I refer to
divergent character displacement as simply (ecological) character
displacement throughout the rest of the text.

\begin{figure}
\centering
\includegraphics{Fig_1_ForagingScenarios}
\caption{\label{fig:foraging_scenarios}\textbf{Ecological foraging
scenarios.} I examined whether the effect of ecological character
displacement on food-web dynamics depended on whether consumers competed
for resources that occurred in the same (a) vs.~different habitats (b).
Note that inferences about character displacement can only be made by
comparing food webs with (right) and without (left) a competing
consumer, so I arbitrarily set \emph{C}\textsubscript{2} = 0 for these
comparisons. The width of each arrow corresponds to the initial attack
rate (\emph{a}\textsubscript{ij}) of consumer \emph{j} on resource
\emph{i}. Note that \emph{C}\textsubscript{1} was pre-adapted to
\emph{R}\textsubscript{1} (\emph{a}\textsubscript{11} \textgreater{}
\emph{a}\textsubscript{21}), while \emph{C}\textsubscript{2} was a
mirror image, being pre-adapted to \emph{R\textsubscript{2}}
(\emph{a}\textsubscript{22} = \emph{a}\textsubscript{11};
\emph{a}\textsubscript{12} = \emph{a}\textsubscript{21}). In each
scenario, I assumed consumer feeding rates increased linearly with
resource abundance. I also relax this assumption and consider a more
realistic functional response when resources occurred in different
habitats (b).}
\end{figure}

\subsection{Food-web dynamics}\label{food-web-dynamics}

Given that character displacement occurred across these foraging
scenarios, I focus here on its consequences for food-web dynamics. To do
this, I analyzed differences in resource abundances and food-web
stability at equilibrium. An equilibrium is reached when there is no
change in the population growth rates of consumers and resources (i.e.,
the rates of change in equation \ref{eq:1} are 0), and solving the
system at this point gives equilibrium abundances for each resource
(\(\hat R_i\)) and each consumer (\(\hat C_j\)). I also compared the
local stability of these food webs using standard methods
\citep{Otto2007}. This stability analysis derives the dominant
eigenvalue, \(\lambda_{\text{max}}\), of the matrix of partial
derivatives of each species' population growth rate (given by equation
\ref{eq:1}) with respect to each species' abundance evaluated at
equilibrium. If \(-\lambda_{max}>0\), then the food web will return to
equilibrium after a small perturbation (i.e., it is locally stable),
with more positive values indicating a faster return time. If
\(-\lambda_{max}<0\), then the food web is not locally stable.

When possible, I derived analytical expressions for the relationship
between consumer attack rates and food-web dynamics. To do this, I
simplified the model by assuming that resources are equivalent
(\(r=r_i\) and \(K=K_i\)) as well as consumers (\(e=e_{ij}\);
\(h=h_{ij}\); \(m=m_j\)), except that consumer attack rates and their
habitat preferences (if present) are mirror images of each other
(\(a_{11}=a_{22}\); \(a_{12}=a_{21}\); \(w_{11}=w_{22}\)). Note that I
arbitrarily set \(C_1\) as being pre-adapted to \(R_1\)
(\(a_{11} > a_{21}\); \(w_{11} > 0.5\)), and therefore \(C_2\) was
pre-adapted to \(R_2\). Controlling for other sources of variability
allowed me to isolate the general effects of character displacement. All
mathematical derivations were conducted in Mathematica
\citep{Mathematica} and are provided in Appendices S1-3.

To gain insight to the eco-evolutionary feedback generated by character
displacement, I conducted simulations using an Adaptive Dynamics
approach. Specifically, after letting consumer and resource abundances
reach a steady state, I created a mutant consumer by randomly choosing
one and modifying its attack rate on one resource by either subtracting
or adding a small constant (0.01 in the following simulations) with
equal probability. The mutant's attack rate on the other resource was
determined by a tradeoff, such that
\(\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1\),
where \(A\) is the total investment in attack rates and \(n\) describes
the shape of the tradeoff \citep{Sargent2006}. This function has the
useful property that it differentiates between cases where intermediate
combinations of \(a_{\text{1},j}\) and \(a_{\text{2},j}\) are higher
than the extremes (when \(n>1\), green line in Fig. \ref{fig:tradeoff})
or, conversely, where the two extremes are higher than intermediate
investments (when \(n<1\), orange line in Fig. \ref{fig:tradeoff}). When
\(n=1\), the tradeoff function is linear, and all combinations of
\(a_{\text{1},j}\) and \(a_{\text{2},j}\) have the same total attack
rate (blue line in Fig. \ref{fig:tradeoff}). Assuming the mutant
consumer was rare, I then determined whether the mutant had higher
relative fitness than the resident consumer, and thus could invade and
replace the resident consumer. If the mutant was able to invade, I
updated the attack rate of the resident consumer to the mutant attack
rate and allowed consumer and resource abundances to reach a steady
state. I then repeated the simulation up to 10,000 times, which was
sufficient for consumers to either reach an evolutionary stable strategy
\citep[ESS,][]{Smith1973} or an evolutionary limit (e.g.,
\(\frac{a_{ij}}{a_{\text{1},j}+a_{\text{2},j}}\) is constrained to a
maximum of 1 and minimum of 0). Unless otherwise noted, I conducted
simulations with the following parameter values: \emph{r} = 1; \emph{K}
= 4; \emph{e} = 0.8; \emph{m} = 1; \emph{A} = 2; \emph{h} = 0.4; and
\(w_{11} = w_{22} = 0.6\). I set an initial value of
\(a_{11} = a_{22} = 1.2\), while \(a_{12}\) and \(a_{21}\) depended on
the value of \emph{n}. I set initial consumer and resource abundances
to: \(R_1 = R_2 = 2\); \(C_1 = C_2 = 1\). All simulations were conducted
in R \citep{R} and the code to reproduce these simulations is publically
available on GitHub (\url{https://github.com/mabarbour/ECD_model}).

\begin{figure}
\centering
\includegraphics{Fig_2_Tradeoffs.pdf}
\caption{\label{fig:tradeoff}\textbf{Evolutionary tradeoffs in consumer
attack rates.} In each foraging scenario, I explored the effects of
three different tradeoffs: intermediate combinations of attack rates
(\emph{a}\textsubscript{1,\emph{j}}, \emph{a}\textsubscript{2,\emph{j}})
are higher than the extremes (green line, \emph{n} \textgreater{} 1);
extreme combinations of attack rates are higher than intermediate
investments (orange line, \emph{n} \textless{} 1); and all combinations
of attack rates have the same total attack rate (blue line, \emph{n} =
1). Points corresponding to attack rates at the beginning of the
simulation for \emph{C}\textsubscript{1}, which was pre-adapted to
\emph{R}\textsubscript{1} (\emph{a}\textsubscript{11} \textgreater{}
\emph{a}\textsubscript{12}). Note that \emph{C}\textsubscript{2} was a
mirror image of \emph{C}\textsubscript{1}, being pre-adapted to
\emph{R}\textsubscript{2} (\emph{a}\textsubscript{22} =
\emph{a}\textsubscript{11}; \emph{a}\textsubscript{12} =
\emph{a}\textsubscript{21}).}
\end{figure}

@@

\subsection{Resources occur in the same
habitat}\label{resources-occur-in-the-same-habitat}

@@

I found that the shape of the tradeoff function qualitatively affects
the relationship between character displacement and resource abundances
in this scenario (Fig. \ref{fig:plot_fig3}a,b). For example, if
consumer's are constrained by a linear tradeoff (blue lines), then there
is no net change in total attack rate (Fig. \ref{fig:plot_fig3}a) and
character displacement has no effect on resource abundances (Fig.
\ref{fig:plot_fig3}b). If the tradeoff is concave down (green lines),
then resource abundances can actually increase under character
displacement (Fig. \ref{fig:plot_fig3}b). This is because the total
attack rate of consumers is maximized at intermediate values
(\(a_{1,j}=a_{2,j}\)) and decreases as consumers diverge (Fig.
\ref{fig:plot_fig3}a). When the tradeoff is concave up (orange lines),
character displacement suppresses resource abundances due to the
increase in total attack rates (Fig. \ref{fig:plot_fig3}a,b). Although
the equation I derived for resource abundances was for the scenario
where both consumers and both resources were present, it accurately
predicts the abundance of resources when a single consumer reaches its
evolutionary stable strategy (ESS; triangles on respective colored lines
in Fig. \ref{fig:plot_fig3}b). This is because a single consumer evolves
to be a generalist that has equal attack rates on each resource
(triangles at 0.5 along x-axis in Fig. \ref{fig:plot_fig3}a), resulting
in equivalent resource abundances.

The effect of character displacement on resources corresponds to its
impact on food-web stability. For example, when character displacement
decreases resource abundances (orange points in Fig.
\ref{fig:plot_fig3}b), there is also a decrease in food-web stability
(Fig. \ref{fig:plot_fig3}c). Character displacement may not affect or
even increase food-web stability (blue and green lines in Fig.
\ref{fig:plot_fig3}c); however, evolution does not favor strong
divergence in these scenarios (blue and green points in Fig.
\ref{fig:plot_fig3}a), which dampens these contingent effects. Note that
the dip in stability occurs when both consumers evolve to be
generalists, a situation that is not favored in any of the foraging
scenarios I examined (Fig. \ref{fig:plot_fig3}c).

@@

This equation implies that if consumers evolve to become specialists on
resources that occur in their preferred habitat (e.g., \(w_{1,j}>0.5\)
and \(a_{1,j}>a_{2,j}\)), then the effective attack rate of consumers
(\(w_{1,j}a_{1,j}+w_{2,j}a_{2,j}\)) will always increase, regardless of
the tradeoff (Fig. \ref{fig:plot_fig3}d). Thus, character displacement
always results in resource suppression (Fig. \ref{fig:plot_fig3}e). Note
that the shape of the tradeoff can modify the effect of character
displacement. This is not so much due to the tradeoff affecting the
magnitude of displacement (it does, but the effect is minor), but
because the form of the tradeoff affects resource abundances when a
single consumer has reached an ESS (triangles in Fig.
\ref{fig:plot_fig3}e). In contrast, resource abundances reach a similar
value when consumers evolve in the presence of a competitor (circles in
Fig. \ref{fig:plot_fig3}e), because character displacement tends to
reach a constraint of complete specialization. It is worth noting that
resource abundances are consistently higher at the single consumer ESS
compared to the predictions I derived for when both consumers are
present (deviation of triangles from respective colored lines in Fig.
\ref{fig:plot_fig3}e). This is because consumers actually evolve
slightly specialized attack rates on the resources that occur in their
non-preferred habitat (deviation of triangles from 0.5 along x-axis in
Fig. \ref{fig:plot_fig3}d).

As seen previously, the effect of character displacement on resource
abundances qualitatively corresponds to its effect on food-web stability
(Fig. \ref{fig:plot_fig3}f). Specifically, character displacement
decreases food-web stability, regardless of the tradeoff in attack
rates. This is not simply a consequence of having an additional consumer
in the system, but emerges from the eco-evolutionary feedback between
character displacement and resource suppression (Fig.
\ref{fig:plot_fig3}c). For example, when the tradeoff is concave up
(orange), the initial two-consumer food web (small circle) is more
stable than when there is only one consumer (small triangle); however,
this pattern switches by the end of the eco-evolutionary simulation
(large points).

\begin{figure}
\centering
\includegraphics{Fig_3_MacArthur_LawlorSmith.pdf}
\caption{\label{fig:plot_fig3}\textbf{Effect of character displacement
on food-web dynamics under different evolutionary tradeoffs and foraging
scenarios.} Lines show predicted values when both consumers and
resources are present. Different line colors correspond to different
tradeoffs in attack rates (green, \emph{n} = 1.15; blue, \emph{n} = 1;
orange, \emph{n} = 0.85). Large circles (two consumers) and triangles
(one consumer) correspond to the end points of the eco-evolutionary
simulation for \emph{C}\textsubscript{1} (the choice to display
\emph{C}\textsubscript{1} was arbitrary), whereas as small shapes
correspond to the starting points (only in stability panels). In both
foraging scenarios, feeding rates increase linearly with resource
abundance, but the equation for the effective attack rate is different.}
\end{figure}

@@

In the third foraging scenario (equation \ref{eq:4}), I observed the
same general effect of character displacement as the previous scenario
(resources in different habitats, but linear functional response). This
is because resource abundances at equilibrium are governed by a similar
dynamic (derived in Appendix S1):

@@

And since evolution favors character displacement toward their preferred
resources (see Appendix S1), the effective attack rate of consumers
(\(w_{1,j}a_{1,j}+w_{2,j}a_{2,j}\)) will always increase, resulting in
lower resource abundances and decreased food-web stability (Appendix S4,
Fig. S1).

@@

\subsection{Resource abundances}\label{resource-abundances}
